---
title: "Reproducible Research: Peer Assessment 1"
author: "jetzgetzlos"
date: "6 avril 2017"
output: html_document
html_document:
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "./figure/")
```

## Reproducible research - Project 1 (week2)

### Loading and preprocessing the data

Unzip the dataset then load the csv file

```{r}
unzip("repdata_data_activity.zip")
activity <- read.csv("activity.csv")
```

Show the first rows and some stats about the dataset
```{r}
head(activity)
str(activity)
```

Convert the column `data` into date format
```{r}
activity$date <- as.Date(strptime(activity$date, format = "%Y-%m-%d"))
str(object = activity) 
```


### What is mean total number of steps taken per day ?
For this part of the assignment, we can ignore the missing values in the dataset.

**Calculate the total number of steps taken per day **
```{r}
library(dplyr, verbose = F, warn.conflicts = FALSE)
activity_by_day <- group_by(activity, date)
average_daily_activity <- summarize(activity_by_day, steps=sum(steps,na.rm=TRUE))
print(average_daily_activity)
```

**Make a histogram of the total number of steps taken each day **

```{r}
library(ggplot2)
g <- ggplot(average_daily_activity, aes(steps)) + geom_histogram(fill = "red", color = "black")
print(g)
```

**Calculate and report the mean and median of the total number of steps taken per day **
```{r}
mean_steps <- mean(average_daily_activity$steps,na.rm=TRUE)
mean_steps
median_steps <- median(average_daily_activity$steps,na.rm=TRUE)
median_steps
#avoir scientific display
options(scipen=999)

```
mean : `r round(mean_steps)`

median : `r round(median_steps)`





### What is the average daily activity pattern?

**Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**

Data preparation
```{r}
activity_by_interval <- group_by(activity, interval)
daily_act_pattern <- summarize(activity_by_interval, av_steps=mean(steps,na.rm=TRUE))
head(daily_act_pattern)
```

```{r}

g <- ggplot(daily_act_pattern, aes(interval, av_steps)) +
    geom_line() +
    xlab("5-minute interval") +
    ylab("average number of step")+
    labs(title = "average number of steps taken averaged across all days")
print(g)
```

**Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**

```{r}
max_interval <- filter(daily_act_pattern,av_steps==max(av_steps))
head(max_interval)
```

The interval num `r max_interval[1,1]` contains the maximum number of steps (`r round(max_interval[1,2])`).


### Imputing missing values

**Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**

```{r}
na_num <- sum(is.na(activity))
na_num
na_pc <- sum(is.na(activity))/nrow(activity)*100
na_pc 
```

There are `r na_num` missing values in the dataset (about `r round(na_pc)` %). It's a significant issue. So, we can't ignore them.

**Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.**

My strategy : Replace the missing values by the mean of steps for the considered interval.




**Create a new dataset that is equal to the original dataset but with the missing data filled in.**
```{r}


replace_by_mean <- function(interval,step){
    
   if (!is.na(step)) {
       return(step)
   } else {
       return (as.numeric(daily_act_pattern[daily_act_pattern$interval == interval,"av_steps"]))
   }
}

activity_filled <- activity
activity_filled$steps <- mapply(replace_by_mean, activity$interval, activity$steps)

```

**Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?**

```{r}
activity_by_day_filled <- group_by(activity_filled, date)
average_daily_activity_filled <- summarize(activity_by_day_filled, steps=sum(steps,na.rm=TRUE))
print(average_daily_activity_filled)
```

```{r}
library(ggplot2)
g <- ggplot(average_daily_activity_filled, aes(steps)) + geom_histogram(fill = "red", color = "black")
print(g)
```

```{r}
mean_steps_n <- mean(average_daily_activity_filled$steps,na.rm=TRUE)
mean_steps_n
median_steps_n <- median(average_daily_activity_filled$steps,na.rm=TRUE)
median_steps_n
options(scipen=999)

```
mean : `r round(mean_steps_n)`      (old value: `r round(mean_steps)`)

median : `r round(median_steps_n)`  (old value:`r round(median_steps)`)

yes, these values differ from the estimates from the first part.

The impact of imputing daly data is to increase the mean and median values. The impact is also to increase the total daily number of steps.





### Are there differences in activity patterns between weekdays and weekends?

Using the filled dataset

**Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.**

First, set the week names in English
```{r}
Sys.setlocale("LC_TIME", "English")
```

```{r}
weekend_or_weekday <- function(date) {
    
    if (weekdays(date) == "Saturday" | weekdays(date) == "Sunday") {
        return("weekend")
    } else {
        return("weekday")
    }
}

activity_filled$day_type <- mapply(weekend_or_weekday, activity_filled$date)
head(activity_filled)
print(activity_filled[7485:7500,])

#Set the factors
activity_filled <- transform(activity_filled, day_type = factor(day_type) )
str(activity_filled)

```



**Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).**

Data Preparation with dplyr library as usual
```{r}
activity_filled_by_interval <- group_by(activity_filled, interval,day_type)
daily_act_pattern_filled <- summarize(activity_filled_by_interval, av_steps=mean(steps,na.rm=TRUE))
head(daily_act_pattern_filled)
```

```{r}
g <- ggplot(daily_act_pattern_filled, aes(interval, av_steps, fill=day_type),) +
    geom_line() +
    facet_grid(day_type ~ .) +
    xlab("5-minute interval") +
    ylab("average number of step")+
    labs(title = "average number of steps taken averaged across all days - weekday and weekend")
print(g)

```