---
author: "jetzgetzlos"
date: "18 avril 2017"
output: html_document
---

```{r setup, include=FALSE, cache = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### _Reproducible research - Project 2 (week4)_

## Analysis of major public health and economic problems caused by severe weather event

### Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

After an exploration and an analysis of the NOAA Storm Database - which report events from 1950 to 2011 - we can conclude that : 

* Tornado is the event type that causes the most public health damage.
* Flood is the event type that causes the most economic damage.


### Data Processing
The data code book is available here : https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

```{r, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
#Unzip the dataset then load the csv file //bunzip2 needs R.utils lib
library(R.utils)
if(!file.exists("repdata_data_StormData.csv")) {
    bunzip2("repdata_data_StormData.csv.bz2","repdata_data_StormData.csv",remove=FALSE)
}

stormData <- read.csv("repdata_data_StormData.csv")

```

There are 902 K observations of 37 variables. It's a file about 550 Mbytes. Loading data takes time.

Show the first rows of the dataset
```{r cache = FALSE}
head(stormData)
```

We need only 7 columns : "EVTYPE", "INJURIES"", "FATALITIES"", "PROPDMG", "PROPDMGEXP", "CROPDMG" and "CROPDMGEXP"
```{r}
storms_light<-stormData[,c(8,23:28)]
rm(stormData)
```





#### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?


Using dplyr lib to group by EVTYPE and sum injuries and fatalities
```{r, message=FALSE, warning=FALSE}
library(dplyr)
```

```{r}
storms_by_evtype <- group_by(storms_light, EVTYPE)
dommages_events <- summarize(storms_by_evtype, 
                             deads =sum(FATALITIES,na.rm=TRUE),
                             injuries=sum(INJURIES,na.rm=TRUE))

dommages_events <- arrange(dommages_events, desc(deads+injuries))

#Save the 10 first rows (10 top harmful economic events)
dommages_events <- dommages_events[1:10,]
dommages_events
```

**Make a bar plot that shows the most harmful events **

Litle reformatting data to create plots : we duplicate rows to split deads and injuries damage, we transform the eventype as factors and we plot the data whith ggplot2 lib.

```{r, echo=TRUE}
deads <- dommages_events[,1:2] 
names(deads) <- c("event", "population_dommage")
injuries <- dommages_events[,1-3]
names(injuries) <- c("event", "population_dommage")
events <- rbind(deads,injuries)
events$type <- c(rep(c("deads"),10), rep(c("injuries"),10))
events <- transform(events, type = factor(type))
print(events)
```

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
```

```{r}
g <- ggplot(events, aes(reorder(event, - population_dommage), population_dommage, fill = type ))   +
    geom_bar(stat = "identity") +
    labs(x = "Event", y = "Number of fatalities/injuries", title= "Population damage") + 
    theme(axis.text.x =  element_text(angle=90))
print(g)
```


**Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health ? : **  According to the plot, Tornado is the event type that causes the most public health damage (both for unjuries and fatalities)








#### Across the United States, which types of events have the greatest economic consequences?

What are the possible values of CROPDMGEXP and PROPDMGEXP columns ?

```{r}
table(storms_light$PROPDMGEXP)
table(storms_light$CROPDMGEXP)
```


In the dateframe, CROPDMGEXP and PROPDMGEXP columns indicate the units :

* H or h = 10^2 (100)
* K or k = 10^3 (1 000)
* M or m = 10^6 (1 000 000)
* B or b = 10^9 (1 000 000 000)

_We choose to condider Other data as NA._

We create a new column with numbers factors from units (= multiplying factors)

```{r}
#Set 2 new columns for the units
storms_light$PROP_unit <- 0
storms_light$CROP_unit <- 0

storms_light$PROP_unit[storms_light$PROPDMGEXP=="H"]<-100
storms_light$PROP_unit[storms_light$PROPDMGEXP=="h"]<-100
storms_light$PROP_unit[storms_light$PROPDMGEXP=="K"]<-1000
storms_light$PROP_unit[storms_light$PROPDMGEXP=="k"]<-1000
storms_light$PROP_unit[storms_light$PROPDMGEXP=="M"]<-1000000
storms_light$PROP_unit[storms_light$PROPDMGEXP=="m"]<-1000000
storms_light$PROP_unit[storms_light$PROPDMGEXP=="B"]<-1000000000

storms_light$CROP_unit[storms_light$CROPDMGEXP=="H"]<-100
storms_light$CROP_unit[storms_light$CROPDMGEXP=="h"]<-100
storms_light$CROP_unit[storms_light$CROPDMGEXP=="K"]<-1000
storms_light$CROP_unit[storms_light$CROPDMGEXP=="k"]<-1000
storms_light$CROP_unit[storms_light$CROPDMGEXP=="M"]<-1000000
storms_light$CROP_unit[storms_light$CROPDMGEXP=="m"]<-1000000
storms_light$CROP_unit[storms_light$CROPDMGEXP=="B"]<-1000000000

table(storms_light$PROP_unit)
table(storms_light$CROP_unit)



```


Using dplyr lib to group by EVTYPE and sum prop and crop damage after applying the multiplying factors

```{r}
storms_by_evtype <- group_by(storms_light, EVTYPE)
dommages_events <- summarize(storms_by_evtype, 
                             prop =sum(PROPDMG*PROP_unit/1000000,na.rm=TRUE),
                             crop=sum(CROPDMG*CROP_unit/1000000,na.rm=TRUE))

dommages_events <- arrange(dommages_events, desc(prop+crop))

#Save the 10 first rows (10 top harmful events)
dommages_events <- dommages_events[1:10,]

dommages_events
```


**Make a bar plot that shows the events with the greatest economic consequences  **

Litle reformatting data to create plots : we duplicate rows to split crop and prop damage, we transform the eventype as factors and we plot the data whith ggplot2 lib.


```{r, echo=TRUE}
prop <- dommages_events[,1:2] 
names(prop) <- c("event", "damage")
crop <- dommages_events[,1-3]
names(crop) <- c("event", "damage")
events <- rbind(prop,crop)
events$type <- c(rep(c("property damage"),10), rep(c("crop damage"),10))
events <- transform(events, type = factor(type))
print(events)
```

```{r}
g <- ggplot(events, aes(reorder(event, - damage), damage, fill = type ))   +
    geom_bar(stat = "identity") +
    labs(x = "Event", y = "Economic damage (million dollars)", title= "Events with the greatest economic consequence") + 
    theme(axis.text.x =  element_text(angle=90))
print(g)
```


**Across the United States, which types of events have the greatest economic consequences? : ** According to the plot, Flood is the event type that causes the most economic damage. Drought is the event type that causes the most crop economic damage.


### Results : 

According to the analysis : 

* Tornado is the event type that causes the most public health damage.
* Flood is the event type that causes the most economic damage. Drought is  the event type that causes the most crop economic damage.

    
   
